---
title: "Sentiment Analysis with Text Mining with R"
author: "Pascal Hermann Kouogang Tafo"
format: html
editor: visual
---

## INTRODUCTION

Human often use their understanding of the emotional intent of words to infer whether a section of text is positive or negative. In Chapter 2 of Text Mining with R,authors introduce sentiment analysis and our assignment consists to reproduce and extend the primary example.

------------------------------------------------------------------------

### PLANNED APPROACH

To tackle this task, we will go work as followed:

-   Load the **tidyverse**, **tidytext**, and **janeaustenr** libraries to mirror the original environment.

-   Process the text of *Emma* and *Pride and Prejudice* using the **bing** and **nrc** lexicons to recreate the net sentiment trajectories.

-   Import a new dataset to test the flexibility of the tidy format.

-   Incorporate a third lexicon to observe how domain-specific emotional tagging differs from general-purpose dictionaries.

-   Visualize the results of all three lexicons against the new corpus to identify shifts in absolute vs. relative sentiment.

------------------------------------------------------------------------

## Step 1: Reproducing the base example Analysis

In this section, we will focus on recreating the original sentiment analysis of Jane Austen's novels using the **bing** lexicon as described by Silge and Robinson

### Let's Install and upload the necessary libraries

```{r setup, message=FALSE}

library(tidyverse)
library(tidytext)
library(janeaustenr)
library(stringr)

```

------------------------------------------------------------------------

### Let's Recreate the net sentiment trajectory for Emma and Pride & Prejudice

In order to reproduce the primary example code from chapter 2 which in our case is the net sentiment trajectory for Emma and Pride & Prejudice, we will use the foundational "tidying" steps required to transform raw text into a format suitable for sentiment analysis using tidy data principles. I use Gemini to help create a Qmd file syntax that runs primary example code from chapter 2.

```{r}


# Tidy the Jane Austen books

tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, regex("^chapter [\\divxlc]", ignore_case = TRUE)))
  ) %>%
  ungroup() %>%
  unnest_tokens(word, text)

head(tidy_books,10)

# Recreate the net sentiment trajectory for Emma and Pride & Prejudice

austen_sentiment <- tidy_books %>%
  filter(book %in% c("Emma", "Pride & Prejudice")) %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(sentiment = positive - negative)

head(austen_sentiment,10)

```

------------------------------------------------------------------------

### Let's Visualize the original trajectory

We will visualize the original trajectory using **the Bing Lexicon**.

```{r}


# Visualize the original trajectory

ggplot(austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x") +
  labs(title = "Sentiment Trajectory in Austen's Novels",
       subtitle = "Using the Bing Lexicon")

```

------------------------------------------------------------------------

## Step 2: Extend Analysis

This second will consist at extending the analysis of the original example conducted above following two specifics ways.

### 1) Let's extend the analysis using a

To successfully reach our goal and text the flexibility of our analysis, we will analyze a collection of NYT news articles by creating a simulation of a small news corpus regarding soccer analytics.

```{r}

# Simulating a small news corpus regarding soccer analytics

NYT_news_data <- tibble(
  article = c(rep("Financial Report", 3), rep("Match Review", 3)),
  text = c(
    "The club reported a disastrous financial quarter with massive losses.",
    "Investors are worried about the debt and the failing market strategy.",
    "The board expressed deep regret over the poor fiscal performance.",
    "The young striker scored a brilliant goal in a spectacular victory.",
    "Fans are delighted with the team's creative and dominant playstyle.",
    "It was an amazing, glorious afternoon for the championship leaders."
  )
)

NYT_news_data

Tidy_news <- NYT_news_data %>%
  group_by(article) %>%
  mutate(line = row_number()) %>%
  ungroup() %>%
  unnest_tokens(word, text)

head(Tidy_news,20)



```

### Interpretation

The resulting **Tidy_news** object is a long-form data frame that breaks down the unstructured news text into individual, analyzable units while maintaining the metadata of which article and line each word came from. This structure allows us to programmatically compare the highly negative vocabulary of the financial reports such as **disastrous** or **losses** against the positive language of the match reviews including **brilliant** or **victory** using **sentiment joins**. Essentially, it prepares the data so that the emotional **sum** of each article type can be calculated and visualized.

------------------------------------------------------------------------

### 2) Let's add the AFINN Lexicon to extend the Analysis.

Here, we will incorporate the AFINN sentiment Lexicon, which provides a numeric ratings between -5 and 5 for each word, allowing a more detailed emotional assessment than the simple "yes/no" or positive/negative categorization found in binary lexicons such as Bing and NRC. Our analysis will then consists of comparing the results on the NYT news articles corpus from the original base example.

**a) Comparison of Bing, NRC, and AFINN on the new corpus**

```{r}

# Let's install  and load the textdata package to download the AFINN lexicon and access this dataset.

#We will install the "textdata" package in our environment to easily render the file.

library(textdata)

# AFINN: Numeric lexicon

news_afinn <- Tidy_news %>%
  inner_join(get_sentiments("afinn")) %>%
  group_by(article) %>%
  summarise(sentiment = sum(value)) %>%
  mutate(method = "AFINN")

# Bing: Binary lexicon

news_bing <- Tidy_news %>%
  inner_join(get_sentiments("bing")) %>%
  count(article, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(sentiment = positive - negative, method = "Bing")

# NRC: Binary/Emotion-based lexicon

news_nrc <- Tidy_news %>%
  inner_join(get_sentiments("nrc") %>% 
               filter(sentiment %in% c("positive", "negative"))) %>%
  count(article, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(sentiment = positive - negative, method = "NRC")

```

**b) Let's bind and Visualize**

We will visualize a lexicon comparison

```{r}

bind_rows(news_afinn, news_bing, news_nrc) %>%
  ggplot(aes(article, sentiment, fill = method)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(title = "Lexicon Comparison on News Corpus",
       y = "Net Sentiment Score")

```

### Interpretation

As observed in the original text, when comparing these tools in text, we see that while they usually agree on whether a story is getting **happier or sadder**, they **measure** that feeling differently. In fact:

-   **AFINN** Shows the highest variance by the scoring method because it gives extra weight to very strong words like **disastrous** or **amazing** and its results show much bigger jumps and drops than the others.

-   **Bing** Often results in lower absolute scores because it is a straightforward **"yes/no"** system that counts every positive word as +1 and every negative word as -1 regardless of the **strength of the word**.

-   **NRC** tends to be biased higher relative to the other two because its dictionary contains a higher percentage of positive words compared to negative ones when compared to the Bing list which is the other binary lexicon.

------------------------------------------------------------------------

## CONCLUSION

By completing this assignment, we can comfortably say that if we were tracking a movie's plot, all three lexicons (**AFINN, Bing, NRC**) would show us the same ups and downs. However, the **AFINN** lexicon would make the peaks look like mountains, while **Bing** would make them look like small hills.
